Analysing Melbourne CBD pedestrian counts data with a focus on imputation methods

### 2017-10-21 thesis_draft/01-chap1.Rmd
update

We would expect that the amount of people passing through the neighbouring sensors will be a strong predictor of the pedestrian counts as they are likely to also pass through. Using geographical neighbours, we can capture rare events effectively. Any large shocks to the counts at the neighbouring sensors will have an effect on the counts to be imputed.

@@

In order to use a mixture of the patterns from two different sensors, we need to use the scaled counts at the neighbouring sensors to avoid the magnitude of the counts having an effect on the predicted counts. We scale the counts by standardising to $\mu = 0, \sigma^2 = 1$. Because the covariance with neighbouring sensors may vary over time, we also add an interaction term between the neighbouring sensor counts and the time of the day.

@@

Firstly, we evaluate the initial model used to impute:
$$\text{where } \quad \mu_{\text{Sensor}, \text{HDay}, \text{Time}} \sim \text{Time}\times \text{HDay} \tag{1}$$

@@

For the purposes of evaluating the imputation method at a location with a small proportion of missing values, we will first look at the counts from Southern Cross Station. At this sensor, the proportion of missing values (adjusted) is `r round(na_prop["Southern Cross Station"]*100,2)`Firstly, we evaluate the goodness of fit of the predictions made within the training period (01/01/2014 to 31/12/2016).

```{r, fig.cap = "A plot of the in-sample Actual vs Predicted pedestrian counts at Southern Cross Station (sensor location with small proportion of missing values) for the simple GLM model and the improved GLM model. The dashed line is $x=y$, representing a reference for perfect fit, while solid line is the linear fit of the points. The improvement is very minor at Southern Cross Station as the data at this sensor is close to complete. Inclusion of month as a predictor is a main source of improvement in the model in this case."}
p1 <- ggplot(data = NULL, aes(x = original_df %>% 
             filter(Date_Time < "2017-01-01") %>%
             .[, "Southern Cross Station"],
           y = fitted_base$`Southern Cross Station`)) +
           geom_point(alpha = 0.05) +
           geom_abline(aes(slope = 1, intercept = 0),
                       colour = viridis(3)[1],
                       alpha = 0.7, size = 1.25,
                       linetype = 2) +
           geom_smooth(method = "lm", size = 1.25, alpha = 0.7, 
                       colour = viridis(3)[3]) +
           theme_minimal() + 
           theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5)) +
           scale_x_continuous(limits = c(0, 4000)) +
           scale_y_continuous(limits = c(0, 4000)) +
           xlab("Actual Pedestrian Count") +
           ylab("Predicted Pedestrian Count /n by simple imputation method")
p2 <- ggplot(data = NULL, aes(x = original_df %>% 
             filter(Date_Time < "2017-01-01") %>%
             .[, "Southern Cross Station"],
           y = fitted_data$`Southern Cross Station`)) +
           geom_point(alpha = 0.05) +
           geom_abline(aes(slope = 1, intercept = 0),
                       colour = viridis(3)[1],
                       alpha = 0.7, size = 1.25,
                       linetype = 2) +
           geom_smooth(method = "lm", size = 1.25, alpha = 0.7, 
                       colour = viridis(3)[3]) +
           theme_minimal() + 
           theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5)) +
           scale_x_continuous(limits = c(0, 4000)) +
           scale_y_continuous(limits = c(0, 4000)) +
           xlab("Actual Pedestrian Count") +
           ylab("Predicted Pedestrian Count \n by improved imputation method")
grid.arrange(p1, p2, nrow = 1)
```

Visually, we can see that the fitted values are good estimates as the relationship between the actuals vs fitted is very close to 1:1. This is seen when we perform a least squares estimate of $\widehat{Fitted}_t = \hat{\beta}_0 + \hat{\beta}_1 Actual_t$, and the estimated $\hat{\beta}_0$ is close to 0 and $\hat{\beta}_1$ is close to 1.

Here we see that the GLM model with only time and date based variables work well for fitting small periods of missing data.

>>>>

The model uses and interaction term between the standardised/scaled counts at the neighbouring sensors and the time of the day (hour factor).

The amount of people passing through the neighbouring sensors will be expected to be a strong predictor of the pedestrian counts as they are likely to also pass through. Using geographical neighbours, it is possible capture rare events effectively as the predictors are also random variables. Any large shocks to the counts at the neighbouring sensors will have an effect on the predicted counts.

@@

In order to use a mixture of the patterns from two different sensors, the use of scaled counts at the neighbouring sensors is required to avoid the magnitude of the counts having an effect on the predicted counts. The counts are scaled by standardising to $\mu = 0, \sigma^2 = 1$ at each sensor (standardised within sensors). Because the covariance with neighbouring sensors may vary over time, an interaction term between the neighbouring sensor counts and the time of the day is included.

@@

Firstly, the initial, simple method of imputation used is evaluated:
$$\text{where } \quad \mu_{\text{Sensor}, \text{HDay}, \text{Time}} \sim \text{Time}\times \text{HDay} \tag{1}$$

@@

For the purposes of evaluating the imputation method at a location with a small proportion of missing values, the counts from Southern Cross Station will be used to evaluated prediction accuracy. At this sensor, the proportion of missing values (adjusted) is `r round(na_prop["Southern Cross Station"]*100,2)`. Firstly, the goodness of fit of the predictions made within the training period (01/01/2014 to 31/12/2016) are evaluated.

```{r, fig.cap = "A plot of the in-sample Actual vs Predicted pedestrian counts at Southern Cross Station (sensor location with small proportion of missing values) for the simple GLM model and the improved GLM model. The dashed line is $x=y$, representing a reference for perfect fit, while solid line is the linear fit of the points. The improvement is very minor at Southern Cross Station as the data at this sensor is close to complete. Inclusion of month as a predictor is a main source of improvement in the model in this case."}
p1 <- ggplot(data = NULL, aes(x = original_df %>% 
             filter(Date_Time < "2017-01-01") %>%
             .[, "Southern Cross Station"],
           y = fitted_base$`Southern Cross Station`)) +
           geom_point(alpha = 0.05) +
           geom_abline(aes(slope = 1, intercept = 0),
                       colour = viridis(3)[1],
                       alpha = 0.7, size = 1.25,
                       linetype = 2) +
           geom_smooth(method = "lm", size = 1.25, alpha = 0.7, 
                       colour = viridis(3)[3]) +
           theme_minimal() + 
           theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5)) +
           scale_x_continuous(limits = c(0, 4000)) +
           scale_y_continuous(limits = c(0, 4000)) +
           xlab("Actual Pedestrian Count") +
           ylab("Predicted Pedestrian Count \n by simple imputation method")
p2 <- ggplot(data = NULL, aes(x = original_df %>% 
             filter(Date_Time < "2017-01-01") %>%
             .[, "Southern Cross Station"],
           y = fitted_data$`Southern Cross Station`)) +
           geom_point(alpha = 0.05) +
           geom_abline(aes(slope = 1, intercept = 0),
                       colour = viridis(3)[1],
                       alpha = 0.7, size = 1.25,
                       linetype = 2) +
           geom_smooth(method = "lm", size = 1.25, alpha = 0.7, 
                       colour = viridis(3)[3]) +
           theme_minimal() + 
           theme(aspect.ratio = 1,
                 plot.title = element_text(hjust = 0.5)) +
           scale_x_continuous(limits = c(0, 4000)) +
           scale_y_continuous(limits = c(0, 4000)) +
           xlab("Actual Pedestrian Count") +
           ylab("Predicted Pedestrian Count \n by improved imputation method")
grid.arrange(p1, p2, nrow = 1)
```

Visually, it can be seen that the fitted values are good estimates of the pedestrian counts as the relationship between the actual values and the fitted values is very close to 1:1. This is best observed when performing a least squares estimate (linear fit) of $\widehat{Fitted}_t = \hat{\beta}_0 + \hat{\beta}_1 Actual_t$, where the estimated $\hat{\beta}_0$ is close to 0 and $\hat{\beta}_1$ is close to 1.

This provides evidence that the GLM model with only time and date based variables work well when there is only small periods of missing data.
